---
title: "isoflavones summary stats"
author: "Maria Sholola"
date: "2023-06-28 and so on"
output: github_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=TRUE) 
```


# Load libraries

```{r}
library(tidyverse) # data wrangling
library(readxl) # read in excel files
library(janitor) # clean up names in dataset
library(corrr) # finding correlations
library(rstatix) # stats
library(ggthemes)
library(ggtext)
library(ggpubr) # for adding pvals etc to plots
library(gridExtra)
library(patchwork)
```
# Isoflavones df

Using df's from data analysis rmd
```{r}
imp_metabind_clust_log2_noQCs <- read_csv("meta_omics_noQCs.csv")
```

Soy isoflavones identified at level 2 ID and up

201.0227_4.086
201.0228_4.555
253.0503_5.048
255.0674_5.078
257.0816_6.056
297.0975_4.072
297.0975_4.498
333.0047_4.33
333.0066_3.893
335.0244_4.36
363.021_3.933
363.0234_4.453
429.0815_4.032
429.082_3.756
429.083_3.201
431.0979_3.907
433.113_4.886
433.1136_4.519
445.0782_4.484
447.092_4.187
447.0928_3.777
449.1074_4.301
459.093_3.801
509.0384_0.699
509.0409_3.291
525.0334_3.606
605.1126_3.023

Significant = p < 0.05, foldchange > 1.6
```{r}
#create df with only significant isoflavone features (level 1 - level 2 ID)
isoflavones <- imp_metabind_clust_log2_noQCs %>%
  dplyr::select(c(1:11), "201.0227_4.086",
"201.0228_4.555",
"253.0503_5.048",
"255.0674_5.078",
"257.0816_6.056",
"297.0975_4.072",
"297.0975_4.498",
"333.0047_4.33",
"333.0066_3.893",
"335.0244_4.36",
"363.021_3.933",
"363.0234_4.453",
"429.0815_4.032",
"429.082_3.756",
"429.083_3.201",
"431.0979_3.907",
"433.113_4.886",
"433.1136_4.519",
"445.0782_4.484",
"447.092_4.187",
"447.0928_3.777",
"449.1074_4.301",
"459.093_3.801",
"497.0695_3.785",
"509.0384_0.699",
"509.0409_3.291",
"525.0334_3.606",
"605.1126_3.023")

isoflavones_tidy <- isoflavones %>%
  pivot_longer(cols = 12:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund_log2")
```


```{r}
# relevel factor columns
isoflavones_tidy$pre_post_intervention <- factor(isoflavones_tidy$pre_post_intervention, levels = c("pre_Yellow", "post_Yellow", "pre_Red", "post_Red"))

isoflavones_tidy$Intervention <- factor(isoflavones_tidy$Intervention,
                                         levels = c("Yellow", "Red"))

# make clean label names
labs_ppintervention <- c("before\ncontrol",
                         "after\ncontrol",
                         "before\nTomato-Soy",
                         "after\nTomato-Soy")

isoflavones_tidy$mz_rt <- as.factor(isoflavones_tidy$mz_rt)

labs_isoflavones <- c("4-Ethylphenol sulfate",
"Ethylphenol sulfate isomer",
"Daidzein",
"Dihydrodaidzein",
"O-DMA",
"Ethylphenol glucuronide isomer 1",
"Ethylphenol glucuronide isomer 2",
"Daidzein sulfate isomer 1",
"Daidzein sulfate isomer 2",
"Dihydrodaidzein sulfate",
"Glycitein sulfate isomer 1",
"Glycitein sulfate isomer 2",
"Daidzein glucuronide isomer 1",
"Daidzein glucuronide isomer 2",
"Daidzein glucuronide isomer 3",
"Dihydrodaidzein glucuronide",
"O-DMA glucuronide isomer 1",
"O-DMA glucuronide isomer 2",
"Genistein glucuronide",
"Dihydrogenistein glucuronide isomer 1",
"Dihydrogenistein glucuronide isomer 2",
"6-OH-O-DMA glucuronide",
"Glycitein glucuronide",
"Equol sulfoglucuronide?",
"Daidzein sulfoglucuronide isomer 1",
"Daidzein sulfoglucuronide isomer 2",
"Genistein sulfoglucuronide",
"Daidzein diglucuronide")

levels(isoflavones_tidy$mz_rt) <- labs_isoflavones
```


# Relative total isoflavone levels boxplots
Soy isoflavones identified from lyc/soy effect list



```{r, fig.width=20, fig.height=20}
my_comparisons <- list( c("pre_Yellow", "post_Yellow"), c("pre_Red", "post_Red"), c("post_Yellow", "post_Red") )



isoflavones_tidy %>% 
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_point() +
  geom_line(aes(group = Subject, colour = as.factor(isoflavones_tidy$Subject)), size = 0.2, show.legend = T) +
  theme_classic(base_size = 18, base_family = "sans") +
  scale_x_discrete(labels = c("pre", "post", "pre", "post")) +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Soy isoflavone metabolites (ID level 2 or higher)",
       subtitle = "FDR adjusted p-values from paired T tests") +
  facet_wrap( ~ mz_rt, scales = "free", ncol = 5) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", paired = TRUE, p.adjust.method = "BH")
```

```{r}
isoflavones_sum <- isoflavones %>%
  mutate(total_isoflavones = rowSums(.[12:ncol(.)]))

isoflavones_sum_tidy <- isoflavones_sum %>%
  pivot_longer(cols = 12:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund_log2")

isoflavones_sum_tidy$pre_post <- factor(isoflavones_sum_tidy$pre_post, levels = c("pre", "post"))

isoflavones_sum_tidy$Intervention <- factor(isoflavones_sum_tidy$Intervention,
                                         levels = c("Yellow", "Red"))

```


```{r, fig.width=12}

isoflavones_sum_tidy %>% 
  filter(mz_rt == "total_isoflavones") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Total isoflavone abundances")


```
## Descriptive stats

### all isoflavones
```{r}
compare_means(rel_abund_log2 ~ pre_post, 
              isoflavones_sum_tidy %>% 
                filter(mz_rt == "total_isoflavones"), 
              method = "t.test", paired = TRUE, group.by = "Intervention")

```



```{r, fig.width=10}
(total_isoflavones_boxplots <- isoflavones_sum_tidy %>%
   filter(mz_rt == "total_isoflavones") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention", facet.by = "Intervention", short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1"),
                    labels = c("Control", "Tomato-Soy")) +
  scale_x_discrete() +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = expression(Log[2]~"Metabolite Abundance"),
       title = "Total Abundance of Isoflavone Metabolites",
       subtitle = "Negative ionization mode"))
```


```{r, eval=FALSE}
ggsave("figures/total_isoflavones.svg", total_isoflavones_boxplots, width = 10)
```


# Putative isoflavones 

```{r}
isoflavones_tidy$pre_post <- factor(isoflavones_tidy$pre_post,
                                         levels = c("pre", "post"))
```




## Ethylphenol sulfate (4- and/or 3-config)
1. isomer 1: 201.0227_4.086 
cluster: 201.0227_4.086;201.0227_4.09

### line graph 
```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "201.0227_4.086") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log 2 relative abundance",
       title = "Possibly ethylphenyl sulfate 1: levels in each patient before/after each intervention")

```

2. isomer 2: 201.0228_4.555
cluster: 1122.0244_4.55;201.0228_4.555;203.0172_4.555;425.0363_4.553
```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "201.0228_4.555") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log 2 relative abundance",
       title = "Possibly ethylphenyl sulfate 2: levels in each patient before/after each intervention")
```


### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "201.0227_4.086") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of 4-ethylphenol sulfate",
       subtitle = "matched with standard") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```

for ASN conference
```{r, fig.width=10}
# make ethylphenyl sulfate isomer 1 df
EPS1_tidy <- isoflavones_tidy %>%
  filter(mz_rt == "201.0227_4.086")

compare_means(rel_abund_log2 ~ pre_post, EPS1_tidy, method = "t.test", paired = TRUE, group.by = "Intervention")

EPS1_tidy$pre_post <- as.factor(EPS1_tidy$pre_post)

EPS1_tidy$pre_post <- factor(EPS1_tidy$pre_post, levels = c("pre", "post"))

(EPS1_bps <- EPS1_tidy %>% 
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention", line.color = "gray", line.size = 1, facet.by = "Intervention", short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1"),
                    labels = c("Control", "Tomato-Soy")) +
  scale_x_discrete() +
  geom_line(aes(group = Subject), colour = "gray", size = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = expression(Log[2]~"Metabolite Abundance"),
       title = "Abundance of 4-Ethylphenyl Sulfate",
       subtitle = "Negative ionization mode, m/z = 201.0227, retention time = 4.08"))
```

confirmed with authentic standard 

```{r, eval=FALSE}
ggsave(filename = "figures/4-EPS.svg", plot = EPS1_bps, width = 12)
```

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "201.0227_4.086") %>%
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) + 
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1")) +
  scale_x_discrete(labels = labs_ppintervention) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly ethylphenyl sulfate 1",
       subtitle = "201.0227_4.086")
```


```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "201.0228_4.555") %>%
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) + 
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1")) +
  scale_x_discrete(labels = labs_ppintervention) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly ethylphenyl sulfate 2",
       subtitle = "201.0228_4.555")
```


```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "201.0228_4.555") %>%
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) + 
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1")) +
  scale_x_discrete(labels = labs_ppintervention) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly ethylphenyl sulfate 2",
       subtitle = "201.0228_4.555")
```


## Ethylphenol glucuronide (4- and/or 3-config)
1. isomer 1: 297.0975_4.498
cluster: 297.0975_4.498;297.1062_4.504;365.0866_4.514

```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "297.0975_4.498") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log 2 relative abundance",
       title = "Possibly ethylphenyl glucuronide 1: levels in each patient before/after each intervention")
```



2. isomer 2: 297.0975_4.072
```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "297.0975_4.072") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly ethylphenyl glucuronide 1: levels in each patient before/after each intervention",
       subtitle = "297.0975_4.072")
```

### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "297.0975_4.498") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \nethylphenol glucuronide isomer",
       subtitle = "297.0975_4.498") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```


```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "297.0975_4.072") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \nethylphenol glucuronide isomer",
       subtitle = "297.0975_4.072") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```




## O-DMA glucuronide

### line graph
isomer 1: 433.113_4.886
```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "433.113_4.886") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly O-DMA glucuronide 1: levels in each patient before/after each intervention",
       subtitle = "433.113_4.886")
```

isomer 2: 433.1136_4.519
```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "433.1136_4.519") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly O-DMA glucuronide 2",
       subtitle = "433.1136_4.519")
```

### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "433.113_4.886") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein O-DMA glucuronide isomer",
       subtitle = "433.113_4.886") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "433.1136_4.519") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein O-DMA glucuronide isomer",
       subtitle = "433.1136_4.519") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```



- Patient 6112: one of their O-dma glucuronides decrease while the other one increases

## O-DMA
257.0816_6.056

### line graph 

```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "257.0816_6.056") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly O-DMA",
       subtitle = "257.0816_6.056")
```

### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "257.0816_6.056") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of O-DMA",
       subtitle = "257.0816_6.056") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```


## Equol monoglucuronide (not significant)
417.1192_4.5

### line graphs

```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "417.1192_4.5") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly equol glucuronide: levels in each patient before/after each intervention",
       subtitle = "417.1192_4.5")
```

## Daidzein sulfate

### line graphs
isomer 1: 333.0047_4.33
```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "333.0047_4.33") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly daidzein sulfate: levels in each patient before/after each intervention",
       subtitle = "333.0047_4.33")
```

isomer 2: 333.0066_3.893
```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "333.0066_3.893") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly daidzein sulfate isomer 2",
       subtitle = "333.0066_3.893")
```

### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "333.0047_4.33") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein sulfate isomer",
       subtitle = "333.0047_4.33") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "333.0066_3.893") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein sulfate isomer",
       subtitle = "333.0066_3.893") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```



## Dihydrodaidzein sulfate
335.0244_4.36

### line graph 

```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "335.0244_4.36") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Possibly dihydrodaizein",
       subtitle = "335.0244_4.36")
```

### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "335.0244_4.36") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndihydrodaidzein sulfate",
       subtitle = "335.0244_4.36") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```

## Daidzein glucuronide

### line graph 
429.083_3.201

```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "429.083_3.201") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Daidzein glucuronide",
       subtitle = "429.083_3.201")
```

### boxplots


```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "429.083_3.201") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein glucuronide isomer",
       subtitle = "429.083_3.201") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```


```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "429.082_3.756") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein glucuronide isomer",
       subtitle = "429.082_3.756") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```


```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "429.0815_4.032") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein glucuronide isomer",
       subtitle = "429.0815_4.032") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```

## Dihydrodaidzein glucuronide

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "431.0979_3.907") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndihydrodaidzein glucuronide",
       subtitle = "431.0979_3.907") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```

## Daidzein

253.0503_5.048

### line graph 

```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "253.0503_5.048") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Daidzein",
       subtitle = "253.0503_5.048")
```

### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "253.0503_5.048") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of Daidzein",
       subtitle = "matched with standard") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```


## Dihydrodaidzein
255.0674_5.078

```{r, fig.width=12}
isoflavones_tidy %>% 
  filter(mz_rt == "255.0674_5.078") %>%
  ggplot(aes(x = pre_post, y = rel_abund_log2, color = Intervention)) +
  geom_point() + 
  geom_line(aes(group = Intervention)) +
  scale_color_manual(values = c("Yellow" = "gold",
                                "Red" = "tomato1")) +
  facet_wrap(vars(Subject), scales = "free_y") + 
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Dihydrodaidzein",
       subtitle = "253.0503_5.048")
```

### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "255.0674_5.078") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of Dihydrodaidzein",
       subtitle = "255.0674_5.078") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```

## Genistein sulfoglucuronide

## Daidzein sulfoglucuronide

### boxplots

```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "509.0409_3.291") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein sulfoglucuronide",
       subtitle = "509.0409_3.291") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```



## Genistein glucuronide

## Glycitein sulfate
(and/or biochanin a sulfate)

## Daidzein diglucuronide

### boxplots


```{r}
isoflavones_tidy %>% 
  filter(mz_rt == "605.1126_3.023") %>%
  ggpaired(x = "pre_post", y = "rel_abund_log2", fill = "Intervention",
           line.color = "gray", 
           line.size = 1, 
           facet.by = "Intervention", 
           short.panel.labs = FALSE, panel.labs = list(Intervention = c("", ""))) +
  scale_fill_manual(values = c("Yellow" = "yellow1",
                               "Red" = "tomato1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_line(aes(group = Subject), colour = "gray", linewidth = 0.15) +
  theme_clean(base_size = 18, base_family = "sans") +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Relative abundance of \ndaidzein diglucuronide",
       subtitle = "605.1126_3.023") +
   stat_compare_means(method = "t.test", paired = TRUE) 
```


# Grouped boxplots

## Read in key 

```{r}
key <- read_xlsx("../annotated-features-table.xlsx",
                 sheet = "Features")
```

```{r}
# use full tidy df
tidy_omics <- imp_metabind_clust_log2_noQCs %>%
  pivot_longer(cols = 12:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund_log2")

# relevel factor columns
tidy_omics$pre_post_intervention <- factor(tidy_omics$pre_post_intervention, levels = c("pre_Yellow", "post_Yellow", "pre_Red", "post_Red"))

tidy_omics$Intervention <- factor(tidy_omics$Intervention,
                                         levels = c("Yellow", "Red"))

# make clean label names
labs_ppintervention <- c("pre\ncontrol",
                         "post\ncontrol",
                         "pre\nTomato-Soy",
                         "post\nTomato-Soy")
```

## Daidzein metabs

```{r}
daidzein_metabs <- key %>%
  filter(`Parent compound` == "Daidzein") %>%
  inner_join(tidy_omics, by = "mz_rt")
```

### Boxplots

#### for me
```{r, fig.width=16, fig.height=12}
daidzein_metabs %>% 
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_point() +
  geom_line(aes(group = Subject, colour = as.factor(daidzein_metabs$Subject)), size = 0.2, show.legend = T) +
  theme_classic(base_size = 18, base_family = "sans") +
  scale_x_discrete(labels = c("pre", "post", "pre", "post")) +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Daidzein metabolites (Level 2 ID or higher)",
       subtitle = "FDR adjusted p-values from paired T tests") +
  facet_wrap( ~ ID, scales = "free", ncol = 5, labeller = labeller(ID = label_wrap_gen(12)))
```

#### cleaner
```{r, fig.width=12, fig.height=20}
my_comparisons <- list( c("pre_Yellow", "post_Yellow"), c("pre_Red", "post_Red"), c("post_Yellow", "post_Red") )

(bps_daidzein_metabs <- daidzein_metabs %>% 
  ggpaired(x = "pre_post_intervention", y = "rel_abund_log2", 
           fill = "Intervention", line.color = "gray",
           id = "Subject", line.size = 0.15) +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1")) +
  geom_point() +
  scale_x_discrete(labels = c("pre", "post", "pre", "post")) +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Daidzein metabolites") +
  facet_wrap( ~ ID, ncol = 3, labeller = labeller(ID = label_wrap_gen(21))) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", paired = TRUE, p.adjust.method = "BH", label = "p.signif") +
  theme_classic(base_size = 18, base_family = "sans") +
  theme(axis.text.x = element_text(angle = 0)) +
  rremove("legend"))
```

```{r}
ggsave("daidzein_metabolites.svg",
       width = 12, height = 20, scale = 1)
```


```{r, fig.height=8}
daidzein_metabs %>% 
  filter(ID %in% c("daidzein glucuronide isomer 1", "daidzein glucuronide isomer 2", "daidzein glucuronide isomer 3")) %>%
  ggpaired(x = "pre_post_intervention", y = "rel_abund_log2", 
           fill = "Intervention", line.color = "gray",
           id = "Subject", line.size = 0.15) +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1"),
                    labels = c("Control", "Tomato-Soy")) +
  geom_point() +
  scale_x_discrete(labels = labs_ppintervention) +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Daidzein glucuronide isomers detected in urine") +
  facet_wrap( ~ ID, ncol = 3, labeller = labeller(ID = label_wrap_gen(12))) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", paired = TRUE, p.adjust.method = "BH", label = "p.signif") +
  theme_classic(base_size = 14, base_family = "sans") +
  theme(axis.text.x = element_text(angle = 45))
```



## Genistein metabs


```{r}
genistein_metabs <- key %>%
  filter(`Parent compound` == "Genistein") %>%
  inner_join(tidy_omics, by = "mz_rt")
```

### Boxplots

#### for me
```{r, fig.width=12, fig.height=12}
genistein_metabs %>% 
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_point() +
  geom_line(aes(group = Subject, colour = as.factor(genistein_metabs$Subject)), size = 0.2, show.legend = T) +
  theme_classic(base_size = 18, base_family = "sans") +
  scale_x_discrete(labels = c("pre", "post", "pre", "post")) +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Genistein metabolites (Level 2 ID or higher)",
       subtitle = "FDR adjusted p-values from paired T tests") +
  facet_wrap( ~ ID, scales = "free", ncol = 3, labeller = labeller(ID = label_wrap_gen(12))) 
```

#### cleaner

```{r, fig.width=8, fig.height=20}
(bps_genistein_metabs <- genistein_metabs %>% 
  ggpaired(x = "pre_post_intervention", y = "rel_abund_log2", 
           fill = "Intervention", line.color = "gray",
           id = "Subject", line.size = 0.15) +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1")) +
  geom_point() +
  scale_x_discrete(labels = c("pre", "post", "pre", "post")) +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Genistein metabolites") +
  facet_wrap( ~ ID, ncol = 3, labeller = labeller(ID = label_wrap_gen(21))) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", paired = TRUE, p.adjust.method = "BH", label = "p.signif") +
  theme_classic(base_size = 18, base_family = "sans") +
  theme(axis.text.x = element_text(angle = 0)) +
  rremove("legend"))
```


```{r}
ggsave("genistein_metabolites.svg",
       width = 8, height = 20, scale = 1)
```

## Glycitein metabs


```{r}
glycitein_metabs <- key %>%
  filter(`Parent compound` == "Glycitein") %>%
  inner_join(tidy_omics, by = "mz_rt")
```

### Boxplots

#### for me
```{r, fig.width=12}
glycitein_metabs %>% 
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1"),
                    labels = c("Control", "Tomato-Soy"),
                    name = "Intervention") +
  geom_point() +
  geom_line(aes(group = Subject, colour = as.factor(glycitein_metabs$Subject)), size = 0.2, show.legend = T) +
  theme_classic(base_size = 18, base_family = "sans") +
  scale_x_discrete(labels = c("pre", "post", "pre", "post")) +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Glycitein metabolites (Level 2 ID or higher)",
       subtitle = "FDR adjusted p-values from paired T tests") +
  facet_wrap( ~ ID, scales = "free", ncol = 3, labeller = labeller(ID = label_wrap_gen(12))) 
```

#### cleaner

```{r, fig.height=20, fig.width=5}
(bps_glycitein_metabs <- glycitein_metabs %>% 
  ggpaired(x = "pre_post_intervention", y = "rel_abund_log2", 
           fill = "Intervention", line.color = "gray",
           id = "Subject", line.size = 0.15) +
  scale_fill_manual(values = c("Red" = "tomato1",
                               "Yellow" = "yellow1"),
                    labels = c("Control", "Tomato-Soy")) +
  geom_point() +
  scale_x_discrete(labels = c("pre", "post", "pre", "post")) +
  labs(x = "",
       y = "Log2 relative abundance",
       title = "Glycitein metabolites") +
  facet_wrap( ~ ID, ncol = 3, labeller = labeller(ID = label_wrap_gen(21))) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", paired = TRUE, p.adjust.method = "BH", label = "p.signif") +
  theme_classic(base_size = 18, base_family = "sans") +
  theme(axis.text.x = element_text(angle = 0)))
```


```{r}
ggsave("glycitein_metabolites.svg",
       width = 5, height = 20, scale = 1)
```

## Plot together

```{r, fig.height=24, fig.width=18}
patchwork_design <- "AABB
                     AABB
                     AABB
                     AABB
                     AACC"

(bps_daidzein_metabs + (bps_genistein_metabs / bps_glycitein_metabs)) +
  plot_layout(design = patchwork_design, guides = "collect") +
  plot_annotation(title = "Soy isoflavone (level 2 ID or higher) levels grouped by parent compound",
                  subtitle = "FDR adjusted p-values from paired T-tests",
                  tag_levels = "A")         
```

```{r}
ggsave("soyisoflavone_metabolites_boxplots.svg",
       width = 22, height = 18, scale = 1)
```

