---
title: "USDA Inflammation Metabolomics Data analysis"
subtitle: "Urine, HILIC (-) LCMS"
author: "Maria Sholola"
date: '2023-06-05'
output: 
  html_document:
    highlight: kate
    theme: yeti
    toc: true
    toc_float: true
    toc_depth: 5
    code_download: true
    fig_width: 7
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=TRUE) 
```

# Load libraries
```{r, warning = FALSE, message = FALSE}
library(readxl) # for reading in excel files
library(janitor) # data checks and cleaning
library(glue) # for easy pasting
library(FactoMineR) # for PCA
library(factoextra) # for PCA
library(rstatix) # for stats
library(pheatmap) # for heatmaps
library(plotly) # for interactive plots
library(htmlwidgets) # for saving interactive plots
library(devtools)
library(notame) # used for feature clustering
library(doParallel)
library(igraph) # feature clustering
library(ggpubr) # visualizations
library(knitr) # clean table printing
library(rmarkdown)
library(mixOmics) # for multilevel PCAs
library(corrr) # correlations
library(ggcorrplot)
library(ggthemes)
library(ggtext)
library(PCAtools)
library(pathview) # for functional analysis and KEGG annotation
library(tidyverse) # for everything
```

# Read in data
```{r}
# raw filtered metabolomics data in HILIC (-)
omicsdata <- read_csv("Feature lists/HILIC-NEG-Filtered-Data-05Jun23_2414features.csv")

# metadata
metadata <- read_excel("Metadata-urine-HILIC-NEG.xlsx")
```

# Wrangle data

```{r}
metadata <- metadata %>%
  rename("sample_ID" = Sample_ID)
```


```{r}
# rename "row ID"
omicsdata <- omicsdata %>%
  rename("row_ID" = `row ID`)

# how many features
nrow(omicsdata)

# are there any duplicates?
omicsdata %>% get_dupes(mz_rt)

```

```{r}
# remove dupes
omicsdata <- omicsdata %>% 
  distinct(mz_rt, .keep_all = TRUE)

# check again for dupes
omicsdata %>% get_dupes(mz_rt)

# how many features
nrow(omicsdata)
```

Sometimes a weird logical column (lgl) comes up in my data. Let's check if it's there

```{r}
colnames(omicsdata)
```


```{r}
# remove weird lgl column
omicsdata <- omicsdata %>%
  dplyr::select(!where(is.logical))

colnames(omicsdata)
```



```{r}
# create long df for omics df
omicsdata_tidy <- omicsdata %>%
  pivot_longer(cols = 3:ncol(.),
               names_to = "sample_ID",
               values_to = "peak_height")

# combine meta and omics dfs
df_combined <- full_join(omicsdata_tidy,
                         metadata,
                         by = c("sample_ID" = "sample_ID"))

# separate mz and rt
df_combined_sep <- df_combined %>%
  separate(col = mz_rt,
           into = c("mz", "rt"),
           sep = "_") 

# convert columns to correct type
df_combined_sep$mz <- as.numeric(df_combined_sep$mz)
df_combined_sep$rt <- as.numeric(df_combined_sep$rt)
df_combined_sep$Subject <- as.character(df_combined_sep$Subject)
df_combined_sep$Intervention <- as.character(df_combined_sep$Intervention)

# rearrange column order
df_combined_sep <- df_combined_sep %>%
 dplyr::select(sample_ID, pre_post, Intervention, everything())

str(df_combined_sep)

# replace NA's in subject and intervention columns with QC
df_combined_sep$Subject <- df_combined_sep$Subject %>%
  replace_na("QC")

df_combined_sep$Intervention <- df_combined_sep$Intervention %>%
  replace_na("QC")


```

# Data summaries

## Number of masses detected
```{r}
nrow(omicsdata)
```


## Mass range for metabolites detected?

```{r}
range(df_combined_sep$mz)
```

## RT range for metabolites detected?

```{r}
range(df_combined_sep$rt)
```

## mass vs RT scatterplot
```{r}
# plot
(plot_mzvsrt <- df_combined_sep %>%
  ggplot(aes(x = rt, y = mz)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Retention time, min",
       y = "m/z, neutral",
       title = "mz across RT for all features"))
```

## Histogram for mass range
```{r}
df_combined_sep %>%
  ggplot(aes(x = mz)) +
  geom_histogram(binwidth = 25) +
  theme_minimal() +
  labs(x = "Monoisotopic mass (amu)",
       y = "Number of features",
       title = "Distribution of features by mass")
```

## Histogram for RT

```{r}
df_combined_sep %>%
  ggplot(aes(x = rt)) +
  geom_histogram(binwidth = 0.1) + # 6 second bins
  theme_minimal() +
  labs(x = "Retention time",
       y = "Number of features",
       title = "Distribution of features by retention time")
```


# NAs and imputing

## NAs
```{r}
# NAs in all data including QCs
NAbyRow <- rowSums(is.na(omicsdata[,-1]))

hist(NAbyRow,
     breaks = 56, # because there are 56 samples, 48 samples + 8 QCs
     xlab = "Number of missing values",
     ylab = "Number of metabolites",
     main = "How many missing values are there?")
```

```{r}
# samples only (no QCs)
omicsdata_noQC <- omicsdata %>%
 dplyr::select(-contains("QC"))

#NAs in samples only?
NAbyRow_noQC <- rowSums(is.na(omicsdata_noQC[,-1]))

hist(NAbyRow_noQC,
     breaks = 48, # because there are 48 samples 
     xlab = "Number of missing values",
     ylab = "Number of metabolites",
     main = "How many missing values are there?")
```

Are there any missing values in QCs? There shouldn't be after data preprocessing/filtering
```{r}
omicsdata_QC <- omicsdata %>%
 dplyr::select(starts_with("P")) 

NAbyRow_QC <- colSums(is.na(omicsdata_QC))
# lets confirm that there are no missing values from my QCs
sum(NAbyRow_QC) # no
```


```{r}
# calculate how many NAs there are per feature in whole data set
contains_NAs <- df_combined %>%
  group_by(mz_rt) %>%
  count(is.na(peak_height)) %>%
  filter(`is.na(peak_height)` == TRUE)
head(contains_NAs)
```




## Remove NAs

To try and handle outliers, I came up with filtering out metabolites that are only present in one person. i.e. remove metabolites that are missing from at least 44 samples. I am taking this bit out for now as it did not change anything

```{r, eval=FALSE}
# remove features that have 44 or more NAs
omit_features <- contains_NAs %>%
  filter(n >= 44)
#preview
nrow(omit_features) # 109 features to remove

# how many features to remove?
nrow(omicsdata) - nrow(omit_features)

# now remove these features from the omics dataset
omicsdata <- omicsdata %>%
  anti_join(omit_features,
            by = "mz_rt")

 # how many features are there now?
nrow(omicsdata)
```


## Data imputation
```{r}
# impute any missing values by replacing them with 1/2 of the lowest peak height value of a feature (i.e. in a row).
imputed_omicsdata <- omicsdata

imputed_omicsdata[] <- lapply(imputed_omicsdata, 
                              function(x) ifelse(is.na(x),
                                                 min(x, na.rm = TRUE)/2, x))

dim(imputed_omicsdata)
```

Are there any NAs?
```{r}
imputed_omicsdata %>%
  is.na() %>%
  sum()

# imputations worked
```


# Create new imputed tidy datasets
```{r}
# create long df for imputed omics df
imputed_omicsdata_tidy <- imputed_omicsdata %>%
  pivot_longer(cols = 3:ncol(.),
               names_to = "sample_ID",
               values_to = "peak_height")

# combine meta and imputed omics dfs
imputed_fulldata <- full_join(imputed_omicsdata_tidy,
                         metadata,
                         by = c("sample_ID" = "sample_ID"))

# separate mz and rt
imputed_fulldata_sep <- imputed_fulldata %>%
  separate(col = mz_rt,
           into = c("mz", "rt"),
           sep = "_") 

# convert columns to correct type
imputed_fulldata_sep$mz <- as.numeric(imputed_fulldata_sep$mz)
imputed_fulldata_sep$rt <- as.numeric(imputed_fulldata_sep$rt)
imputed_fulldata_sep$Subject <- as.character(imputed_fulldata_sep$Subject)
imputed_fulldata_sep$Intervention <- as.character(imputed_fulldata_sep$Intervention)
```


# Notame feature reduction

vignette for reference
```{r}
#browseVignettes("notame")
```

## Plot features. RT vs mz before notame
```{r}
# rt vs mz plot
imputed_fulldata_sep %>%
  ggplot(aes(x = rt, y = mz)) +
  geom_point() +
  theme_minimal() +
  labs(x = "RT (min)",
       y = "mz")
```


## Data restructuring for notame
```{r}
# create features list from imputed data set to only include unique feature ID's (mz_rt), mz and RT
features <- imputed_fulldata_sep %>%
  cbind(imputed_fulldata$mz_rt) %>%
  rename("mz_rt" = "imputed_fulldata$mz_rt") %>%
 dplyr::select(c(mz_rt, mz, rt)) %>%
  distinct() # remove the duplicate rows

# create a second data frame which is just imputed_fulldata restructured to another wide format
data_notame <- data.frame(imputed_omicsdata %>%
                           dplyr::select(-row_ID) %>%
                            t())

data_notame <- data_notame %>%
  tibble::rownames_to_column() %>% # change samples from rownames to its own column
  row_to_names(row_number = 1) # change the feature IDs (mz_rt) from first row obs into column name
```

Check structures
```{r}
# check if mz and rt are numeric
str(features)
tibble(features)
```

```{r}
# check if results are numeric
tibble(data_notame)

# change to results to numeric
data_notame <- data_notame %>%
  mutate_at(-1, as.numeric)

tibble(data_notame)
```


## Find connections
```{r}
connection <- find_connections(data = data_notame,
                               features = features,
                               corr_thresh = 0.9,
                               rt_window = 1/60,
                               name_col = "mz_rt",
                               mz_col = "mz",
                               rt_col = "rt")

head(connection)
```

## Clustering
```{r}
clusters <- find_clusters(connections = connection, d_thresh = 0.8)
```

```{r}
# assign a cluster ID to all features. Clusters are named after feature with highest median peak height
features_clustered <- assign_cluster_id(data_notame, clusters, features, name_col = "mz_rt")

# visualize clusters
#visualize_clusters(data_notame, features, clusters, min_size = 3, rt_window = 2,name_col = "mz_rt", mz_col = "mz", rt_col = "rt", file_path = "~/path/to/project/")

# lets see how many features are removed when we only keep one feature per cluster
pulled <- pull_clusters(data_notame, features_clustered, name_col = "mz_rt")
cluster_data <- pulled$cdata




cluster_features <- pulled$cfeatures

nrow(omicsdata) - nrow(cluster_features)
```

```{r}
# export clustered feature list
write_csv(cluster_features,
          "cluster-features_hilic-neg.csv")
```


 ## Reduce dataset based on clustering
```{r}
# transpose the full dataset back to wide so that it is more similar to the notame dataset
imputed_fulldata_wide <- imputed_fulldata %>%
 dplyr::select(-"row_ID") %>%
  pivot_wider(names_from = mz_rt,
              values_from = peak_height)

# list of reduced features
clusternames <- cluster_features$mz_rt

#dplyr::select only the features are in the reduced list
imp_clust <- imputed_fulldata_wide[,c(names(imputed_fulldata_wide) %in% clusternames)]

# bind back sample names
imp_clust <- cbind(imputed_fulldata_wide[1], imp_clust)

tibble(imp_clust)

```

## Mz vs RT scatterplot 

```{r}
# plot new rt vs mz scatterplot post-clustering
(plot_mzvsrt_postcluster <- cluster_features %>%
  ggplot(aes(x = rt,
             y = mz)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Retention time, min",
       y = "m/z, neutral",
       title = "mz across RT for all features after clustering"))


```


```{r}
# plot both scatterplots to compare with and without notame clustering
(scatterplots <- ggarrange(plot_mzvsrt, 
                           plot_mzvsrt_postcluster, 
                           nrow = 2))
```

# Bind meta data
```{r}
imp_metabind_clust <- right_join(metadata, 
                                 imp_clust,
                                 by = "sample_ID")
```

# Visualize untransformed data

## Data wrangling
```{r}
# change meta data columns to character so that I can change NAs from QCs to "QC"
imp_metabind_clust <- imp_metabind_clust %>%
  mutate_at(c("Subject",
              "Period",
              "Intervention",
              "pre_post",
              "sequence",
              "Intervention_week",
              "Sex",
              "Age",
              "BMI"),
            as.character) 

# replace NAs in metadata columns for QCs
imp_metabind_clust[is.na(imp_metabind_clust)] <- "QC"

# unite pre_post column with intervention column to create pre_intervention column
imp_metabind_clust <- imp_metabind_clust %>%
  unite(col = "pre_post_intervention",
        c("pre_post","Intervention"),
        sep = "_",
        remove = FALSE)

# long df
imp_metabind_clust_tidy <- imp_metabind_clust %>%
  pivot_longer(cols = 12:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund")

# structure check
str(imp_metabind_clust_tidy)
```



## Boxplot
```{r}
imp_metabind_clust_tidy %>%
  ggplot(aes(x = sample_ID, y = rel_abund, color = Intervention)) +
  geom_boxplot(alpha = 0.6) +
  scale_color_manual(values = c("light grey", "tomato1", "gold")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "LC-MS (-) Feature Abundances by Sample",
       subtitle = "Unscaled data",
       y = "Relative abundance")
```
Will need to log transform in order to normalize and actually see the data

# Log2 transform
```{r}
imp_metabind_clust_tidy_log2 <- imp_metabind_clust_tidy %>%
  mutate(rel_abund_log2 = if_else(rel_abund > 0, log2(rel_abund), 0)) %>%
  replace(is.na(.), 0)
```

## Boxplot
```{r}
(bp_data_quality <- imp_metabind_clust_tidy_log2 %>%
  mutate(Sample = if_else(Intervention %in% c("Yellow", "Red"), "Sample", "QC")) %>%
  ggplot(aes(x = sample_ID, y = rel_abund_log2, fill = Sample))) +
  geom_boxplot(alpha = 0.6) +
  scale_fill_manual(values = c("light gray", "deepskyblue2")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "HILIC (-) Feature Abundances by Sample",
       subtitle = "Log2 transformed data",
       y = "Relative abundance")
```

Much better! QCs look great.

# Load Feature ID Key
Here, I'm inserting a key that would indicate identified features above level 2 ID.

```{r}
key_omics <- read_xlsx("../annotated-features-table.xlsx",
                 sheet = "Features") %>%
  clean_names() %>%
  filter(lc_mode_c18_hilic == "HILIC") %>% # hilic
  filter(esi_mode == "-") %>% # ESI -
  dplyr::select(mz_rt, id, metabolite_class, parent_compound) # select relevant columns

# add key columns (left_join to only keep all observations present in full feature table)
anno_imp_metabind_clust_tidy_log2 <- left_join(imp_metabind_clust_tidy_log2, key_omics, by = "mz_rt") %>%
  # replace NAs in feature ID columns to un-annotated feature id's (mz_rt)
  mutate(Feature_ID = coalesce(id, mz_rt))
```

# PCAs

## With QCS

### Wrangle

```{r}
# go back to wide data
anno_imp_metabind_clust_log2 <- anno_imp_metabind_clust_tidy_log2 %>%
  select(c(1:11),
         Feature_ID, rel_abund_log2) %>%
  pivot_wider(names_from = Feature_ID,
              values_from = rel_abund_log2)

PCA.imp_metabind_clust_log2 <- PCA(anno_imp_metabind_clust_log2,  # wide data
                                   quali.sup = 1:11, # remove qualitative variables
                                   graph = FALSE, # don't graph
                                   scale.unit = FALSE) # don't scale, already transformed data

# PCA summary
kable(summary(PCA.imp_metabind_clust_log2))
```

```{r}
# pull PC coordinates into df
PC_coord_QC_log2 <- as.data.frame(PCA.imp_metabind_clust_log2$ind$coord)

# bind back metadata from cols 1-10
PC_coord_QC_log2 <- bind_cols(anno_imp_metabind_clust_log2[,1:11], PC_coord_QC_log2)

# grab some variance explained
importance_QC <- PCA.imp_metabind_clust_log2$eig

# set variance explained with PC1, round to 2 digits
PC1_withQC <- round(importance_QC[1,2], 2)

# set variance explained with PC2, round to 2 digits
PC2_withQC <- round(importance_QC[2,2], 2)
```

### Plots
Using FactoExtra package
```{r}
# scree plot
fviz_eig(PCA.imp_metabind_clust_log2)

# get eigenvalues
kable(get_eig(PCA.imp_metabind_clust_log2))
```

```{r}
# scores plot
fviz_pca_ind(PCA.imp_metabind_clust_log2)
```

```{r}
# loadings
fviz_pca_var(PCA.imp_metabind_clust_log2)
```

### Manual scores plots
```{r}
# manual scores plot
(PCA_withQCs <- PC_coord_QC_log2 %>%
  ggplot(aes(x = Dim.1, y = Dim.2,
             fill = factor(Intervention, levels = c("Yellow", "Red", "QC")))) +
  geom_point(shape = 21, alpha = 0.8) +
  scale_fill_manual(values = c("gold", "tomato1", "light grey")) +
  scale_color_manual(values = "black") +  
  theme_minimal() +
  coord_fixed(PC2_withQC/PC1_withQC) +
  labs(x = glue::glue("PC1: {PC1_withQC}%"),
       y = glue::glue("PC2: {PC2_withQC}%"),
       fill = "Group",
       title = "Principal Components Analysis Scores Plot",
       subtitle = "Log2 transformed data"))
```

## Without QCs

### Wrangle 

```{r}
anno_imp_metabind_clust_log2_noQCs <- anno_imp_metabind_clust_log2 %>%
  filter(Intervention != "QC")

PCA.imp_metabind_clust_log2_noQCs <- PCA(anno_imp_metabind_clust_log2_noQCs, # wide data
                               quali.sup=1:11, # remove qualitative variables
                               graph=FALSE, # don't graph
                               scale.unit=FALSE) # don't scale, we already did this

# look at summary
kable(summary(PCA.imp_metabind_clust_log2_noQCs))
```

```{r}
# pull PC coordinates into df
PC_coord_noQCs_log2 <- as.data.frame(PCA.imp_metabind_clust_log2_noQCs$ind$coord)

# bind back metadata from cols 1-10
PC_coord_noQCs_log2 <- bind_cols(anno_imp_metabind_clust_log2_noQCs[,1:11], PC_coord_noQCs_log2)

# grab some variance explained
importance_noQC <- PCA.imp_metabind_clust_log2_noQCs$eig

# set variance explained with PC1, round to 2 digits
PC1_noQC <- round(importance_noQC[1,2], 2)

# set variance explained with PC2, round to 2 digits
PC2_noQC <- round(importance_noQC[2,2], 2)
```

### Plots
Using FactoExtra

```{r}
# scree plot
fviz_eig(PCA.imp_metabind_clust_log2_noQCs)
```

```{r}
# scores plot
fviz_pca_ind(PCA.imp_metabind_clust_log2_noQCs)
```


```{r}
# plot of contributions from features to PC1
(var_contrib_noQCs_PC1 <- fviz_contrib(PCA.imp_metabind_clust_log2_noQCs,
             choice = "var",
             axes = 1,
             top = 25,
             title = "Var contribution to PC1: no QCs"))

# plot of contributions from features to PC2
(var_contrib_noQCs_PC2 <- fviz_contrib(PCA.imp_metabind_clust_log2_noQCs,
             choice = "var",
             axes = 2,
             top = 25,
             title = "Var contribution to PC2: no QCs"))
```


```{r}
# loadings
fviz_pca_var(PCA.imp_metabind_clust_log2_noQCs) # nightmare
```
### Manual scores plots

#### Yellow vs red
```{r}
(PCA_withoutQCs <- PC_coord_noQCs_log2 %>%
  ggplot(aes(x = Dim.1, 
             y = Dim.2,
             fill = Intervention)) +
    geom_point(shape = 21, alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha=0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha=0.5) +
    scale_fill_manual(values = c("gold", "tomato1")) +
    scale_color_manual(values = "black") +  
    theme_minimal() +
    coord_fixed(PC2_noQC/PC1_noQC) +
    labs(x = glue::glue("PC1: {PC1_noQC}%"),
         y = glue::glue("PC2: {PC2_noQC}%"),
         fill = "Intervention",
         title = "Principal Components Analysis Scores Plot",
         subtitle = "Log2 transformed data, without QCs"))
ggplotly(PCA_withoutQCs)
```

#### pre vs post
```{r}
(PCA_withoutQCs.pre_post <- PC_coord_noQCs_log2 %>%
  ggplot(aes(x = Dim.1, 
             y = Dim.2,
             fill = factor(pre_post_intervention, levels = c("pre_Yellow",
                                                             "post_Yellow",
                                                             "pre_Red",
                                                             "post_Red")),
             text = sample_ID)) +
    geom_point(shape = 21, alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha=0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha=0.5) +
    scale_fill_manual(values = c("gray", "yellow1", "pink1", "red2")) +
    scale_color_manual(values = "black") +  
    theme_minimal() +
    coord_fixed(PC2_noQC/PC1_noQC) +
    labs(x = glue::glue("PC1: {PC1_noQC}%"),
         y = glue::glue("PC2: {PC2_noQC}%"),
         fill = "pre_post",
         title = "Principal Components Analysis Scores Plot",
         subtitle = "Log2 transformed, without QCs"))
ggplotly(PCA_withoutQCs.pre_post,
         tooltip = "text") 
```

## Notes 

Looks like subject 6106 is an outlier. Let's take a look at what the data looks like without the outlier.


## Removal of outlier

### With QCs

#### Wrangle

```{r}
# go back to wide data
imp_no6106_log2 <- anno_imp_metabind_clust_log2 %>%
  filter(Subject != 6106)

PCA.imp_no6106_log2 <- PCA(imp_no6106_log2,  # wide data
                                   quali.sup = 1:11, # remove qualitative variables
                                   graph = FALSE, # don't graph
                                   scale.unit = FALSE) # don't scale, already transformed data

# PCA summary
summary(PCA.imp_no6106_log2)
```


```{r}
# pull PC coordinates into df
PC_no6106_QC_log2 <- as.data.frame(PCA.imp_no6106_log2$ind$coord)

# bind back metadata from cols 1-11
PC_no6106_QC_log2 <- bind_cols(imp_no6106_log2[,1:11], PC_no6106_QC_log2)

# grab some variance explained
importance_no6106_QC <- PCA.imp_no6106_log2$eig

# set variance explained with PC1, round to 2 digits
PC1_no6106_withQC <- round(importance_no6106_QC[1,2], 2)

# set variance explained with PC2, round to 2 digits
PC2_no6106_withQC <- round(importance_no6106_QC[2,2], 2)
```

#### Plots
Using FactoExtra package
```{r}
# scree plot
fviz_eig(PCA.imp_no6106_log2)
```

```{r}
# scores plot
fviz_pca_ind(PCA.imp_no6106_log2)
```

#### Manual scores plots

##### Pre vs post
```{r}
(PCA_no6106_withQCs_prepost <- PC_no6106_QC_log2 %>%
  ggplot(aes(x = Dim.1, 
             y = Dim.2,
             fill = factor(pre_post_intervention, levels = c("pre_Yellow",
                                                             "post_Yellow",
                                                             "pre_Red",
                                                             "post_Red")),
             text = sample_ID)) +
    geom_point(shape = 21, alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha=0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha=0.5) +
    scale_fill_manual(values = c("gray", "yellow1", "pink1", "red2")) +
    scale_color_manual(values = "black") +  
    theme_minimal() +
    coord_fixed(PC2_no6106_withQC/PC1_no6106_withQC) +
    labs(x = glue::glue("PC1: {PC1_no6106_withQC}%"),
         y = glue::glue("PC2: {PC2_no6106_withQC}%"),
         fill = "pre_post",
         title = "Principal Components Analysis Scores Plot"))
ggplotly(PCA_no6106_withQCs_prepost,
         tooltip = "text") 
```


### Without QCs

#### Wrangle
```{r}
imp_no6106_log2_noQCs <- imp_no6106_log2 %>%
  filter(Intervention != "QC")

PCA.imp_no6106_log2_noQCs <- PCA(imp_no6106_log2_noQCs, # wide data
                               quali.sup=1:11, # remove qualitative variables
                               graph=FALSE, # don't graph
                               scale.unit=FALSE) # don't scale, we already did this

# look at summary
kable(summary(PCA.imp_no6106_log2_noQCs))
```

```{r}
# pull PC coordinates into df
PC_coord_no6106_noQC_log2 <- as.data.frame(PCA.imp_no6106_log2_noQCs$ind$coord)

# bind back metadata from cols 1-11
PC_coord_no6106_noQC_log2 <- bind_cols(imp_no6106_log2_noQCs[,1:11], PC_coord_no6106_noQC_log2)

# grab some variance explained
importance_no6106_noQC <- PCA.imp_no6106_log2_noQCs$eig

# set variance explained with PC1, round to 2 digits
PC1_no6106_noQC <- round(importance_no6106_noQC[1,2], 2)

# set variance explained with PC2, round to 2 digits
PC2_no6106_noQC <- round(importance_no6106_noQC[2,2], 2)
```


#### Plots
Using FactoExtra
```{r}
# scree plot
fviz_eig(PCA.imp_no6106_log2_noQCs)
```

```{r}
# scores plot
fviz_pca_ind(PCA.imp_no6106_log2_noQCs)
```


#### Manual scores plots

##### Pre vs post
```{r}
(PCA_no6106_withoutQCs.prepost_intervention <- PC_coord_no6106_noQC_log2 %>%
  ggplot(aes(x = Dim.1, 
             y = Dim.2,
             fill = factor(pre_post_intervention, levels = c("pre_Yellow",
                                                             "post_Yellow",
                                                             "pre_Red",
                                                             "post_Red")),
             text = sample_ID)) +
    geom_point(shape = 21, alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha=0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha=0.5) +
    scale_fill_manual(values = c("gray", "yellow1", "pink1", "red2")) +
    scale_color_manual(values = "black") +  
    theme_minimal() +
    coord_fixed(PC2_no6106_noQC/PC1_no6106_noQC) +
    labs(x = glue::glue("PC1: {PC1_no6106_noQC}%"),
         y = glue::glue("PC2: {PC2_no6106_noQC}%"),
         fill = "pre_post",
         title = "Principal Components Analysis Scores Plot",
         subtitle = "Log2 transformed data, no 6106"))
ggplotly(PCA_no6106_withoutQCs.prepost_intervention,
         tooltip = "text") 
```



# PCAtools

## No outliers

### Data wrangling

```{r}
# create rel abund df suitable for PCAtools package

imp_clust_omicsdata_forPCAtools <- as.data.frame(t(anno_imp_metabind_clust_log2[,-c(2:11)])) # transpose df 

names(imp_clust_omicsdata_forPCAtools) <- imp_clust_omicsdata_forPCAtools[1,] # make sample IDs column names

imp_clust_omicsdata_forPCAtools <- imp_clust_omicsdata_forPCAtools[-1,] # remove sample ID row

imp_clust_omicsdata_forPCAtools <- imp_clust_omicsdata_forPCAtools %>%
  dplyr::select(!contains("QC")) # remove QC observations


# create metadata df suitable for PCAtools pckg
metadata_forPCAtools <- metadata %>%
  column_to_rownames(var = "sample_ID") # change sample ID column to rownames

# create a vector so that col names in abundance df matches metadata df
order_forPCAtools <- match(rownames(metadata_forPCAtools), colnames(imp_clust_omicsdata_forPCAtools))

# reorder col names in abundance df so that it matches metadata
abundata_reordered_forPCAtools <- imp_clust_omicsdata_forPCAtools[ ,order_forPCAtools] 

# change abundance df to numeric
log2_abundata_reordered_forPCAtools <- abundata_reordered_forPCAtools %>%
  mutate_at(1:ncol(.), as.numeric)

# fix rownames to have keys again
rownames(log2_abundata_reordered_forPCAtools) <- rownames(abundata_reordered_forPCAtools)

# remove 6106 subj from both df
log2_abundata_reordered_forPCAtools <- log2_abundata_reordered_forPCAtools %>%
  dplyr::select(!contains("6106"))

metadata_forPCAtools <- metadata_forPCAtools %>%
  filter(Subject != 6106)

# unite pre_post column with intervention column to create pre_intervention column
metadata_forPCAtools <- metadata_forPCAtools %>%
  unite(col = "pre_post_intervention",
        c("pre_post","Intervention"),
        sep = "_",
        remove = FALSE)

```

### PCA
```{r, fig.width=7, fig.height=7.5}
# pca
p <- pca(log2_abundata_reordered_forPCAtools, 
         metadata = metadata_forPCAtools, 
         scale = FALSE # using scaled data already (log2 transformed)
         
)

# plot

PCAtools::biplot(p,
                 showLoadings = TRUE, # show variables that contribute the most to PCs
                 lab = NULL,
                 title = )
```

### Screeplot analysis

Horn's parallel analysis
```{r, warning=FALSE}
horn <- parallelPCA(log2_abundata_reordered_forPCAtools)

horn$n
```

Elbow method
```{r}
elbow <- findElbowPoint(p$variance)

elbow
```

```{r, fig.width=7, fig.height=7.5}

  screeplot(p,
    components = getComponents(p, 1:20),
    vline = c(horn$n, elbow)) +
  geom_label(aes(x = horn$n + 1, y = 50,
      label = 'Horn\'s', vjust = -1, size = 8)) +
    geom_label(aes(x = elbow + 1, y = 50,
      label = 'Elbow method', vjust = -3, size = 8))
```

How many PCs do we need to capture at least 80% variance?
```{r}
which(cumsum(p$variance) > 80)[1]
```

### More PCAs

#### Pre vs post both

##### PC1vPC2


```{r, fig.width=10, fig.height=10}
biplot(p,
       lab = paste0(metadata_forPCAtools$Subject),
       colby = 'pre_post_intervention',
      colkey = c("pre_Yellow" = "lemonchiffon1",
                                        "post_Yellow" = "yellow2",
                                        "pre_Red" = "rosybrown2",
                                        "post_Red" = "darkred"),
       hline = 0, vline = 0,
       # ellipse config
       ellipse = TRUE,
       ellipseType = 't', # assumes multivariate t-distribution
       ellipseLevel = 0.99,
       ellipseFill = TRUE,
       ellipseAlpha = 0.2,
       ellipseLineSize = 0,
       xlim = c(-80,80), ylim = c(-40,40),
       legendPosition = 'right',
       title = "PCA Scores Plot",
       subtitle = "Log2 transformed data, HILIC (-), subject 6106 removed, no QCs \n99% confidence level ellipses")

```


```{r, fig.width=8.5, fig.height=6.5}
(PCA.colby.prevspost <- biplot(p,
                               lab = NULL,
                           colby = 'pre_post_intervention',
                           colkey = c("pre_Yellow" = "lemonchiffon1",
                                        "post_Yellow" = "yellow2",
                                        "pre_Red" = "rosybrown2",
                                        "post_Red" = "darkred"),
                           hline = 0, vline = 0,
         legendPosition = 'right',
         title = "PCA Scores Plot",
         subtitle = "Log2 transformed data, HILIC (-), subject 6106 removed, no QCs",
         showLoadings = TRUE))
```


##### Pairs plot
```{r, fig.width=10, fig.height=10}
(PCA_pairsplot.colby.prevspost <-
  pairsplot(p,
    components = getComponents(p, c(1:10)),
    triangle = TRUE, trianglelabSize = 12,
    hline = 0, vline = 0,
    pointSize = 0.4,
    gridlines.major = FALSE, gridlines.minor = FALSE,
    colby = 'pre_post_intervention', 
    colkey = c("pre_Yellow" = "yellow",
               "post_Yellow" = "yellow4",
               "pre_Red" = "pink",
               "post_Red" = "red4"),
    title = 'Pairs plot', plotaxes = FALSE,
    margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm')))

```

### Sex

#### PC1vPC2
```{r, fig.width=7, fig.height=7.5}
(PCA.colby.Sex <- biplot(p,
                           lab = paste0(metadata_forPCAtools$Subject),
                          colby = 'Sex',
                          colkey = c("M" = "red",
                                     "F" = "purple"),
                          hline = 0, vline = 0,
                          legendPosition = 'right' +
                            geom_point(aes(text = metadata_forPCAtools$Subject))))

ggplotly(PCA.colby.Sex,
         tooltip = "text") 

```

#### Pairsplot
```{r, fig.width=10, fig.height=10}
  pairsplot(p,
    components = getComponents(p, c(1:10)),
    triangle = TRUE, trianglelabSize = 12,
    hline = 0, vline = 0,
    pointSize = 0.4,
    gridlines.major = FALSE, gridlines.minor = FALSE,
    colby = 'Sex', 
    colkey = c("M" = "red",
               "F" = "purple"),
    title = 'Pairs plot', plotaxes = FALSE,
    margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
```


### Overall pre v post

#### PC1vPC2
```{r, fig.width=7, fig.height=7.5}
(PCA.colby.overall.prevspost <- biplot(p,
                                       lab = paste0(metadata_forPCAtools$Subject),
                                       colby = 'pre_post',
                                       colkey = c("pre" = "orange",
                                                  "post" = "green3"),
                                       hline = 0, vline = 0,
                                       legendPosition = 'right' +
                                         geom_point(aes(text = metadata_forPCAtools$Subject))))

ggplotly(PCA.colby.overall.prevspost,
         tooltip = "text") 

```


```{r, fig.width=10, fig.height=10}
  pairsplot(p,
    components = getComponents(p, c(1:10)),
    triangle = TRUE, trianglelabSize = 12,
    hline = 0, vline = 0,
    pointSize = 0.4,
    gridlines.major = FALSE, gridlines.minor = FALSE,
    colby = 'pre_post', 
    colkey = c("pre" = "orange",
               "post" = "green3"),
    title = 'Pairs plot', plotaxes = FALSE,
    margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
```


### Period

#### PC1vPC2
```{r, fig.width=7, fig.height=7.5}
(PCA.colby.period <- biplot(p,
                            lab = paste0(metadata_forPCAtools$Subject),
                            colby = 'Period',
                            hline = 0, vline = 0,
                            legendPosition = 'right' +
                              geom_point(aes(text = metadata_forPCAtools$Subject))))

ggplotly(PCA.colby.period,
         tooltip = "text") 

```

#### Pairsplot
```{r, fig.width=10, fig.height=10}
  pairsplot(p,
    components = getComponents(p, c(1:10)),
    triangle = TRUE, trianglelabSize = 12,
    hline = 0, vline = 0,
    pointSize = 0.4,
    gridlines.major = FALSE, gridlines.minor = FALSE,
    colby = 'Period',
    title = 'Pairs plot', plotaxes = FALSE,
    margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
```



### Sequence

#### PC1vPC2
```{r, fig.width=7, fig.height=7.5}
(PCA.colby.sequence <- biplot(p,
                            lab = paste0(metadata_forPCAtools$Subject),
                            colby = 'sequence',
                            hline = 0, vline = 0,
                            legendPosition = 'right' +
                              geom_point(aes(text = metadata_forPCAtools$Subject))))

ggplotly(PCA.colby.sequence,
         tooltip = "text") 

```

#### Pairsplot
```{r, fig.width=10, fig.height=10}
  pairsplot(p,
    components = getComponents(p, c(1:10)),
    triangle = TRUE, trianglelabSize = 12,
    hline = 0, vline = 0,
    pointSize = 0.4,
    gridlines.major = FALSE, gridlines.minor = FALSE,
    colby = 'sequence',
    title = 'Pairs plot', plotaxes = FALSE,
    margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
```





## Eigen corplots
This is a cool way to explore the correlations between the metadata and the PCs! I want to look at how the metavariables correlate with PCs that account for 80% variation in the dataset. 

Again: How many PCs do we need to capture at least 80% variance?
```{r}
which(cumsum(p$variance) > 80)[1]
```

```{r, fig.width=12, fig.height=7.5}

  eigencorplot(p,
    components = getComponents(p, 1:18), # get components that account for 80% variance
    metavars = colnames(metadata_forPCAtools),
    col = c('darkblue', 'blue2', 'gray', 'red2', 'darkred'),
    cexCorval = 0.7,
    colCorval = 'white',
    fontCorval = 2,
    posLab = 'bottomleft',
    rotLabX = 45,
    posColKey = 'top',
    cexLabColKey = 1.5,
    scale = TRUE,
    main = 'PC1-18 metadata correlations',
    colFrame = 'white',
    plotRsquared = FALSE)


```


```{r, fig.width=14, fig.height=7.5}
  eigencorplot(p,
    components = getComponents(p, 1:18),
    metavars = colnames(metadata_forPCAtools),
    col = c('white', 'cornsilk1', 'gold', 'forestgreen', 'darkgreen'),
    cexCorval = 1.2,
    fontCorval = 2,
    posLab = 'all',
    rotLabX = 45,
    scale = TRUE,
    main = bquote(Principal ~ component ~ Pearson ~ r^2 ~ metadata ~ correlates),
    plotRsquared = TRUE,
    corFUN = 'pearson',
    corUSE = 'pairwise.complete.obs',
    corMultipleTestCorrection = 'BH',
    signifSymbols = c('****', '***', '**', '*', ''),
    signifCutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1))
```


# Univariate analysis

## Wrangle data
```{r}
# use tidy data 
head(anno_imp_metabind_clust_tidy_log2)

# remove QCs
df_for_stats <- anno_imp_metabind_clust_tidy_log2 %>%
  filter(Intervention != "QC")

# check if QCs were removed
unique(df_for_stats$Intervention)
```


```{r}
# df without outlier 6106
df_for_stats_noOutlier <- df_for_stats %>%
  filter(Subject != "6106")

# check if outlier was removed
unique(df_for_stats_noOutlier$Subject)
```


```{r}
# turn off sci notation outputs
options(scipen = 999)
```


## Parametric tests

## ANOVA (repeated measures) across timepoints

```{r}
anova_outpout_df <- df_for_stats_noOutlier %>% 
  dplyr::select(Subject, pre_post_intervention, Feature_ID, rel_abund_log2) %>%
  group_by(Feature_ID) %>%
  anova_test(rel_abund_log2 ~ pre_post_intervention, wid = Subject,
             detailed = TRUE) %>%
  adjust_pvalue(method = "BH") %>%
  as.data.frame()

anova_sig <- anova_outpout_df %>%
  filter(p.adj <= .05)

# how many significant features?
nrow(anova_sig)
  
```

```{r}
# tukey's posthoc
tukey_anova <- df_for_stats_noOutlier %>% 
  dplyr::select(Subject, pre_post_intervention, Feature_ID, rel_abund_log2) %>%
  group_by(Feature_ID) %>%
  tukey_hsd(rel_abund_log2 ~ pre_post_intervention, wid = subject)
```


### Heatmap of features significant by ANOVA

```{r, fig.height=20, fig.width=12, fig.asp=1}
# go back to wide for stats df (keeping all subjects in the visualization, but anova was done without outliers)
df_for_stats_wide <- df_for_stats %>%
  dplyr::select(!c(rel_abund, metabolite_class, parent_compound, id, mz_rt)) %>%
  pivot_wider(names_from = Feature_ID,
              values_from = rel_abund_log2)

ANOVA_heatmap_data <- df_for_stats_wide %>%
  unite("Subject_pre_post_intervention", Subject, pre_post_intervention, sep = "_", remove = FALSE) %>%
  dplyr::select(Subject, Subject_pre_post_intervention, pre_post, all_of(anova_sig$Feature_ID))


ANOVA_heatmap_annotated <- 
  pheatmap(t(ANOVA_heatmap_data[,-c(1:3)]),
           scale = "row",
           cluster_rows = TRUE,
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           clustering_method = "ward.D2",
           labels_col = ANOVA_heatmap_data$Subject_pre_post_intervention,
           color = colorRampPalette(c("#67a9cf", "#f7f7f7", "#ef8a62"))(16),
           main = "Heatmap of features significant across timepoints \nby repeated measures one-way ANOVA \nBenjamoni-Hochberg corrected p-values > 0.05 \nHILIC (-)")
  
```



```{r}
ANOVA_heatmap_data2 <- df_for_stats_wide %>%
  filter(Subject != 6106) %>%
  unite("Subject_pre_post_intervention", Subject, pre_post_intervention, sep = "_", remove = FALSE) %>%
  dplyr::select(sample_ID, Subject_pre_post_intervention, pre_post_intervention, all_of(anova_sig$Feature_ID)) %>%
  column_to_rownames("sample_ID") 
```


```{r}
# change pre_post_intervention to factor
df_for_stats_wide$pre_post_intervention <- as.factor(df_for_stats_wide$pre_post_intervention)

# create annotation rows for treatment and wrangle
# select sample col from heatmap metadata (also ensures the order is correct)
anno_trt_row <- as.data.frame(rownames(ANOVA_heatmap_data2))

# pull desired columns
anno_trt_row$pre_post_intervention <- ANOVA_heatmap_data2$pre_post_intervention

# select trt
anno_trt_row <- anno_trt_row %>%
  dplyr::select(pre_post_intervention) 

# get rownames to match heatmap
rownames(anno_trt_row) <- rownames(ANOVA_heatmap_data2)

# create annotation colors
annotation_colors <- list(pre_post_intervention = c("pre_Yellow" = "lemonchiffon1",
                                                    "post_Yellow" = "yellow2",
                                                    "pre_Red" = "rosybrown2",
                                                    "post_Red" = "darkred"))
```

```{r, fig.height=20, fig.width=12, fig.asp=1}
ANOVA_heatmap <- 
  pheatmap(t(ANOVA_heatmap_data2[,-c(1:2)]),
           scale = "row",
           cluster_rows = TRUE,
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           clustering_method = "ward.D2",
           labels_col = ANOVA_heatmap_data2$Subject_pre_post_intervention,
           color = colorRampPalette(c("#67a9cf", "#f7f7f7", "#ef8a62"))(16),
           annotation_col = anno_trt_row,
           annotation_colors = annotation_colors,
           annotation_names_col = F,
           main = "Heatmap of features significant across timepoints by repeated measures one-way ANOVA \nHILIC-MS (-)", fontfamily = "sans")
  
```



```{r}
ggsave(plot = ANOVA_heatmap, height = 10, width = 12,
       filename = "plots and figures/ANOVA_sig_heatmap.svg")
```


### Paired t tests

Here, I am comparing pre- to post-intervention for both yellow and tomato soy (Red) juice interventions. I also compare post-yellow to post-red intervention. I am using the log transformed values of rel abundance since parametric tests assume normality.

#### Ctrl
```{r}
# run paired t-tests for control intervention
ctrl_t.test_paired <- df_for_stats %>%
  filter(Intervention == "Yellow") %>%
  dplyr::select(Subject, pre_post, Feature_ID, rel_abund_log2) %>%
  group_by(Feature_ID) %>%
  t_test(rel_abund_log2 ~ pre_post, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
ctrl_t.test_paired_sig <- ctrl_t.test_paired %>%
  filter(p <= 0.05)
kable(ctrl_t.test_paired_sig)

# how many are significant?
nrow(ctrl_t.test_paired_sig)
```


#### Ctrl - no outlier
```{r}
# run paired t-tests for control intervention
ctrl_noOutlier_t.test_paired <- df_for_stats_noOutlier %>%
  filter(Intervention == "Yellow") %>%
  dplyr::select(Subject, pre_post, mz_rt, Feature_ID, rel_abund_log2) %>%
  group_by(Feature_ID) %>%
  t_test(rel_abund_log2 ~ pre_post, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
ctrl_noOutlier_t.test_paired_sig <- ctrl_noOutlier_t.test_paired %>%
  filter(p <= 0.05)
kable(ctrl_noOutlier_t.test_paired_sig)

# how many are significant?
nrow(ctrl_noOutlier_t.test_paired_sig)
```


#### Red
```{r}
# run paired t-tests for control intervention
red_t.test_paired <- df_for_stats %>%
  filter(Intervention == "Red") %>%
  dplyr::select(Subject, pre_post, mz_rt, Feature_ID, rel_abund_log2) %>%
  group_by(Feature_ID) %>%
  t_test(rel_abund_log2 ~ pre_post, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
red_t.test_paired_sig <- red_t.test_paired %>%
  filter(p <= 0.05)
kable(red_t.test_paired_sig)

# how many are significant?
nrow(red_t.test_paired_sig)
```


#### Red Without outlier
```{r}
# run paired t-tests for control intervention
red_noOutlier_t.test_paired <- df_for_stats_noOutlier %>%
  filter(Intervention == "Red") %>%
  dplyr::select(Subject, pre_post, mz_rt, Feature_ID, rel_abund_log2) %>%
  group_by(Feature_ID) %>%
  t_test(rel_abund_log2 ~ pre_post, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
red_noOutlier_t.test_paired_sig <- red_noOutlier_t.test_paired %>%
  filter(p <= 0.05)
kable(red_noOutlier_t.test_paired_sig)

# how many are significant?
nrow(red_noOutlier_t.test_paired_sig)
```


#### Post-red vs post-yellow

```{r}
# run paired t-tests for post interventions
post_t.test_paired <- df_for_stats %>%
  filter(pre_post == "post") %>%
  dplyr::select(Subject, Intervention, mz_rt, Feature_ID, rel_abund_log2) %>%
  group_by(Feature_ID) %>%
  t_test(rel_abund_log2 ~ Intervention, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
post_t.test_paired_sig <- post_t.test_paired %>%
  filter(p <= 0.05)
kable(post_t.test_paired_sig)

# how many are significant?
nrow(post_t.test_paired_sig)
```

#### Post-red vs post-yellow Without Outlier

```{r}
# run paired t-tests for post interventions
post_noOutlier_t.test_paired <- df_for_stats %>%
  filter(pre_post == "post",
         Subject != "6106") %>%
  dplyr::select(Subject, Intervention, mz_rt, Feature_ID, rel_abund_log2) %>%
  group_by(Feature_ID) %>%
  t_test(rel_abund_log2 ~ Intervention, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
post_noOutlier_t.test_paired_sig <- post_noOutlier_t.test_paired %>%
  filter(p <= 0.05)
kable(post_noOutlier_t.test_paired_sig)

# how many are significant?
nrow(post_noOutlier_t.test_paired_sig)
```


### Paired t tests

Here, I am comparing pre- to post-intervention for both yellow and tomato soy (Red) juice interventions. I also compare post-yellow to post-red intervention. I am using the log transformed values of rel abundance since parametric tests assume normality.

#### Ctrl
```{r}
# run paired t-tests for control intervention
ctrl_t.test_paired <- df_for_stats %>%
  filter(Intervention == "Yellow") %>%
 dplyr::select(Subject, pre_post, mz_rt, rel_abund_log2) %>%
  group_by(mz_rt) %>%
  t_test(rel_abund_log2 ~ pre_post, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
ctrl_t.test_paired_sig <- ctrl_t.test_paired %>%
  filter(p <= 0.05)
tibble(ctrl_t.test_paired_sig)

# how many are significant?
nrow(ctrl_t.test_paired_sig)
```


#### Ctrl Without outlier
```{r}
# run paired t-tests for control intervention
ctrl_noOutlier_t.test_paired <- df_for_stats_noOutlier %>%
  filter(Intervention == "Yellow") %>%
 dplyr::select(Subject, pre_post, mz_rt, rel_abund_log2) %>%
  group_by(mz_rt) %>%
  t_test(rel_abund_log2 ~ pre_post, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
ctrl_noOutlier_t.test_paired_sig <- ctrl_noOutlier_t.test_paired %>%
  filter(p <= 0.05)
tibble(ctrl_noOutlier_t.test_paired_sig)

# how many are significant?
nrow(ctrl_noOutlier_t.test_paired_sig)
```



#### Red
```{r}
# run paired t-tests for control intervention
red_t.test_paired <- df_for_stats %>%
  filter(Intervention == "Red") %>%
 dplyr::select(Subject, pre_post, mz_rt, rel_abund_log2) %>%
  group_by(mz_rt) %>%
  t_test(rel_abund_log2 ~ pre_post, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
red_t.test_paired_sig <- red_t.test_paired %>%
  filter(p <= 0.05)
tibble(red_t.test_paired_sig)

# how many are significant?
nrow(red_t.test_paired_sig)
```


#### Red Without outlier
```{r}
# run paired t-tests for control intervention
red_noOutlier_t.test_paired <- df_for_stats_noOutlier %>%
  filter(Intervention == "Red") %>%
 dplyr::select(Subject, pre_post, mz_rt, rel_abund_log2) %>%
  group_by(mz_rt) %>%
  t_test(rel_abund_log2 ~ pre_post, 
         paired = TRUE, 
         p.adjust.method = "BH") %>% # Benjamini-Hochberg controlling to lower false positives
  add_significance()
```

Statistically significant features
```{r}
# which features are significant?
red_noOutlier_t.test_paired_sig <- red_noOutlier_t.test_paired %>%
  filter(p <= 0.05)
tibble(red_noOutlier_t.test_paired_sig)

# how many are significant?
nrow(red_noOutlier_t.test_paired_sig)
```

### Volcano plots

#### Post-intervention comparisons 

##### Wrangle (no outlier)
```{r}
# calculate mean rel abund (not log) by sample, and avg fold change (FC) diff by feature
red_v_yel_volcano_data_noOutlier <- df_for_stats %>%
  filter(pre_post == "post",
         Subject != 6106) %>% # remove outlier subj
  group_by(Intervention, mz_rt) %>%
  summarize(mean_rel_abund = mean(rel_abund)) %>%
  pivot_wider(names_from = Intervention, values_from = mean_rel_abund) %>%
  mutate(FC_postRed_div_postYellow = Red/Yellow) 

# bind back pval
red_v_yel_tobind_noOutlier <- post_noOutlier_t.test_paired %>%
  dplyr::select(p)

# calculate log2FC, and -log10p
red_v_yel_volcano_data_noOutlier <- 
  bind_cols(red_v_yel_volcano_data_noOutlier, red_v_yel_tobind_noOutlier) %>%
  mutate(log2_FC_postRed_div_postYellow = if_else(FC_postRed_div_postYellow > 0,
                                                  log2(FC_postRed_div_postYellow),
                                                  -(log2(abs(FC_postRed_div_postYellow)))), 
         neglog10p = -log10(p))


# create a df of features passing FC and pval cutoffs higher in post-Red
postRed_sig_red_v_yel_volcano_noOutlier <- red_v_yel_volcano_data_noOutlier %>%
  filter(p <= 0.05 & log2_FC_postRed_div_postYellow >= 0.6)

# create a df of features passing FC and pval cutoffs higher in post-control
postYellow_sig_red_v_yel_volcano_noOutlier <- red_v_yel_volcano_data_noOutlier %>%
  filter(p <= 0.05 & log2_FC_postRed_div_postYellow <= -0.6)  
```

##### Export sig features

Create feature lists with significant features along with their clusters
```{r}
#post-Red list
postRed_sig_intrvntn_comp_clusters <- left_join(postRed_sig_red_v_yel_volcano_noOutlier, cluster_features,
                                               by = "mz_rt")

write_csv(postRed_sig_intrvntn_comp_clusters,
          "Feature lists/postRed-sigfeatures-intervntn-comp.csv")

#post-Yellow list
postYellow_sig_intrvntn_comp_clusters <- left_join(postYellow_sig_red_v_yel_volcano_noOutlier, cluster_features,
                                               by = "mz_rt")

write_csv(postYellow_sig_intrvntn_comp_clusters,
          "Feature lists/postYellow-sigfeatures-intervntn-comp.csv")
```




##### Plot
```{r}
(red_v_yellow_volcano_noOutlier <- red_v_yel_volcano_data_noOutlier %>%
  ggplot(aes(x = log2_FC_postRed_div_postYellow, y = neglog10p, 
             text = glue("Mass_retention time: {mz_rt}
                         P-value: {p}
                         Fold change tomato/control: {round(FC_postRed_div_postYellow, 2)}"))) +
  geom_point(color = "grey") +
  geom_point(data = postRed_sig_intrvntn_comp_clusters, 
             aes(x = log2_FC_postRed_div_postYellow, y = neglog10p),
             color = "tomato") +
  geom_point(data = postYellow_sig_intrvntn_comp_clusters, 
             aes(x = log2_FC_postRed_div_postYellow, y = neglog10p),
             color = "yellow2") +
  geom_vline(xintercept = 0.6, linetype = "dashed", color = "grey") +
  geom_vline(xintercept = -0.6, linetype = "dashed", color = "grey") +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "grey") +
  coord_cartesian(xlim = c(-5, 8)) +
  labs(title = "Volcano Plot of Features Different in People After Tomato-Soy and Control Juice Consumption",
       subtitle = "Red points are higher after tomato-soy juice consumption while yellow points are higher \nafter control tomato juice consumption. Subject 6106 removed",
       caption = "Vertical dashed lines represent a log2 fold change > 0.6 or < -0.6, and horizontal dashed line represents an FDR corrected \np-value of 0.05.",
       x = "Log2 Fold Change (TomatoSoy/Control)",
       y = "-Log10(P-value)") +
  theme_bw() +
  theme(plot.title = element_text(size = 12, hjust = 0),
        plot.caption = element_text(hjust = 0.5)))

(red_v_yellow_volcano_ggplotly_noOutlier <- ggplotly(red_v_yellow_volcano_noOutlier, tooltip = "text"))
```

Save plots
```{r, eval=FALSE}
# save volcano plot
ggsave(plot = red_v_yellow_volcano_noOutlier,
       filename = "plots and figures/volcano plots/red_v_yellow_volcano_noOutlier.svg")

# save interactive volcano plot
saveWidget(widget = red_v_yellow_volcano_ggplotly_noOutlier,
           file = "plots and figures/volcano plots/interactive_redvyellow_volcano_plot_noOutlier.html")
```



#### Red

##### Wrangle (no outlier)
```{r}
# calculate mean rel abund (not log) by sample, and avg fold change (FC) diff by feature
red_volcano_data_noOutlier <- df_for_stats %>%
  filter(Intervention == "Red",
         Subject != 6106) %>%
  group_by(pre_post, mz_rt) %>%
  summarize(mean_rel_abund = mean(rel_abund)) %>%
  pivot_wider(names_from = pre_post, values_from = mean_rel_abund) %>%
  mutate(FC_post_div_pre = post/pre) 

# bind back pval
red_tobind_noOutlier <- red_noOutlier_t.test_paired %>%
  dplyr::select(p)

# calculate log2FC, and -log10p
red_volcano_data_noOutlier <- 
  bind_cols(red_volcano_data_noOutlier, red_tobind_noOutlier) %>%
  mutate(log2_FC_post_div_pre = log2(FC_post_div_pre),
         neglog10p = -log10(p))


# create a df of features passing FC and pval cutoffs higher in post-intervention compared to pre
red_pre_v_post_volcano_noOutlier <- red_volcano_data_noOutlier %>%
  filter(p <= 0.05 & abs(log2_FC_post_div_pre) >= 0.6)

```

##### Export sig features

Create feature lists with significant features along with their clusters
```{r}
red_sig_prepost_comp_clusters <- left_join(red_pre_v_post_volcano_noOutlier, cluster_features,
                                               by = "mz_rt")

write_csv(red_sig_prepost_comp_clusters,
          "Feature lists/Red-sigfeatures-PrevsPost-noOutliers.csv")

```


##### Plot
```{r}
(red_volcano_noOutlier <- red_volcano_data_noOutlier %>%
  ggplot(aes(x = log2_FC_post_div_pre, y = neglog10p, 
             text = glue("Mass_retention time: {mz_rt}
                         P-value: {p}
                         Fold change post/pre: {round(FC_post_div_pre, 2)}"))) +
  geom_point(color = "grey") +
  geom_point(data = red_sig_prepost_comp_clusters, 
             aes(x = log2_FC_post_div_pre, y = neglog10p),
             color = "tomato") +
  geom_vline(xintercept = 0.6, linetype = "dashed", color = "grey") +
  geom_vline(xintercept = -0.6, linetype = "dashed", color = "grey") + 
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "grey") +
  coord_cartesian(xlim = c(-5, 6)) +
  labs(title = "Volcano Plot of Features Different in People After Tomato-Soy Juice Consumption",
       subtitle = "Red points are higher after tomato-soy juice consumption when compared to prior to consumption",
       caption = "Vertical dashed lines represent an abs(log2 fold change) > 0.6, and horizontal dashed line represents an FDR corrected \np-value of 0.05.",
       x = "Log2 Fold Change (Post/Pre)",
       y = "-Log10(P-value)") +
  theme_bw() +
  theme(plot.title = element_text(size = 12, hjust = 0),
        plot.caption = element_text(hjust = 0.5)))

(red_volcano_ggplotly_noOutlier <- ggplotly(red_volcano_noOutlier, tooltip = "text"))
```

Save plots
```{r, eval=FALSE}
# save volcano plot
ggsave(plot = red_volcano_noOutlier,
       filename = "plots and figures/volcano plots/red_volcano_noOutlier.svg")

# save interactive volcano plot
saveWidget(widget = red_volcano_ggplotly_noOutlier,
           file = "plots and figures/volcano plots/interactive_red_volcano_plot_noOutlier.html")
```

#### Yellow

##### Wrangle (no outlier)
```{r}
# calculate mean rel abund (not log) by sample, and avg fold change (FC) diff by feature
yel_volcano_data_noOutlier <- df_for_stats %>%
  filter(Intervention == "Yellow",
         Subject != 6106) %>%
  group_by(pre_post, mz_rt) %>%
  summarize(mean_rel_abund = mean(rel_abund)) %>%
  pivot_wider(names_from = pre_post, values_from = mean_rel_abund) %>%
  mutate(FC_post_div_pre = post/pre) 

# bind back pval
yel_tobind_noOutlier <- ctrl_noOutlier_t.test_paired %>%
  dplyr::select(p)

# calculate log2FC, and -log10p
yel_volcano_data_noOutlier <- 
  bind_cols(yel_volcano_data_noOutlier, yel_tobind_noOutlier) %>%
  mutate(log2_FC_post_div_pre = log2(FC_post_div_pre),
         neglog10p = -log10(p))


# create a df of features passing FC and pval cutoffs higher in post-intervention compared to pre
yel_pre_v_post_volcano_noOutlier <- yel_volcano_data_noOutlier %>%
  filter(p <= 0.05 & abs(log2_FC_post_div_pre) >= 0.6)

```

##### Export sig features

Create feature lists with significant features along with their clusters
```{r}
yel_sig_prepost_comp_clusters <- left_join(yel_pre_v_post_volcano_noOutlier, cluster_features,
                                               by = "mz_rt")

write_csv(yel_sig_prepost_comp_clusters,
          "Yellow-sigfeatures-PrevsPost-noOutliers.csv")

```

##### Plot
```{r}
(yel_volcano_noOutlier <- yel_volcano_data_noOutlier %>%
  ggplot(aes(x = log2_FC_post_div_pre, y = neglog10p, 
             text = glue("Mass_retention time: {mz_rt}
                         P-value: {p}
                         Fold change post/pre: {round(FC_post_div_pre, 2)}"))) +
  geom_point(color = "grey") +
  geom_point(data = yel_sig_prepost_comp_clusters, 
             aes(x = log2_FC_post_div_pre, y = neglog10p),
             color = "yellow2") +
  geom_vline(xintercept = 0.6, linetype = "dashed", color = "grey") +
  geom_vline(xintercept = -0.6, linetype = "dashed", color = "grey") +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "grey") +
  coord_cartesian(xlim = c(-6, 6)) +
  labs(title = "Volcano Plot of Features Different in People After Control, Yellow Tomato Juice Consumption",
       subtitle = "Yellow points are higher after control juice consumption when compared to prior to consumption.\nSubject 6106 removed",
       caption = "Vertical dashed lines represent abs(log2 fold change) > 0.6, and horizontal dashed line represents an FDR corrected \np-value of 0.05.",
       x = "Log2 Fold Change (Post/Pre)",
       y = "-Log10(P-value)") +
  theme_bw() +
  theme(plot.title = element_text(size = 12, hjust = 0),
        plot.caption = element_text(hjust = 0.5)))

(yel_volcano_ggplotly_noOutlier <- ggplotly(yel_volcano_noOutlier, tooltip = "text"))
```

Save plots
```{r, eval=FALSE}
# save volcano plot
ggsave(plot = yel_volcano_noOutlier,
       filename = "plots and figures/volcano plots/yel_volcano_noOutlier.svg")

# save interactive volcano plot
saveWidget(widget = yel_volcano_ggplotly_noOutlier,
           file = "plots and figures/volcano plots/interactive_yel_volcano_plot_noOutlier.html")
```


## Joined lists
Here, I want to venn significant features before I begin to look more into them. I am interested in the following effects: tomato effect, lycopene and/or soy isoflavones effect.

### Tomato effect
1. Tomato effect: join a list that only keeps features that are both significant in pre vs. post-red and pre vs. post-yellow. Manually deleted the features that don't have the same directionality

```{r}
# keep only features present in both pre vs post red and pre vs post yellow
tomato_effect <- inner_join(red_sig_prepost_comp_clusters,
                            yel_sig_prepost_comp_clusters,
                            by = "mz_rt")
dim(tomato_effect)
```

Export venned list
```{r}
write_csv(tomato_effect,
          "Feature lists/sig-features-tomato-effect.csv")
```

#### Boxplots


##### all sig tomato features

```{r}
# metabs with pval < 0.05 and fc >= 1.51
sigmetabs_tomato_effect <- tomato_effect$mz_rt 
```

```{r}
tomato_effect_df_for_stats_wide <- df_for_stats %>%
  dplyr::select(!rel_abund) %>%
  pivot_wider(names_from = mz_rt,
              values_from = rel_abund_log2) %>%
  dplyr::select(c(1:11),
                sigmetabs_tomato_effect)

# make tidy df
tomato_effect_df_for_stats_tidy <- tomato_effect_df_for_stats_wide %>%
  pivot_longer(cols = 12:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund_log2")
```

```{r}
# changing factor levels for pre_post_intervention
tomato_effect_df_for_stats_tidy$pre_post_intervention <- factor(tomato_effect_df_for_stats_tidy$pre_post_intervention,
                              levels = c("pre_Yellow", "post_Yellow", "pre_Red", "post_Red"))

levels(tomato_effect_df_for_stats_tidy$pre_post_intervention)  
```

```{r, fig.height=15, fig.width=12}
tomato_effect_df_for_stats_tidy %>% 
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = c("Yellow" = "gold",
                                           "Red" = "tomato1")) +
  geom_line(aes(group = Subject), fill = "brown", linewidth = 0.2) +
  facet_wrap(vars(mz_rt), scales = "free_y") + 
  theme_clean() 
```

##### dopamine sulfate

Looking at the mass associated with dopamine sulfate. There are 2 retention times, one of them is a significant feature for tomato effect. The other feature (same mass, diff RT) is the only one I collected MS/MS data for. So I am looking to see if the trends for both features are the same, and the likelihood that they are configurational isomers. 

```{r}
dopamineSO4H_df_for_stats_wide <- df_for_stats_noOutlier %>%
  dplyr::select(!rel_abund) %>%
  pivot_wider(names_from = mz_rt,
              values_from = rel_abund_log2) %>%
  dplyr::select(c(1:11),
                "232.0284_4.836",
                "232.0283_5.427")

# make tidy df
dopamineSO4H_df_forstats_tidy <- dopamineSO4H_df_for_stats_wide %>%
  pivot_longer(cols = 12:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund_log2")
```


```{r}
# changing factor levels for pre_post_intervention
dopamineSO4H_df_forstats_tidy$pre_post_intervention <- factor(dopamineSO4H_df_forstats_tidy$pre_post_intervention,
                              levels = c("pre_Yellow", "post_Yellow", "pre_Red", "post_Red"))

levels(dopamineSO4H_df_forstats_tidy$pre_post_intervention)  
```

```{r}
dopamineSO4H_df_forstats_tidy %>% 
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = c("Yellow" = "gold",
                                           "Red" = "tomato1")) +
  geom_line(aes(group = Subject), fill = "brown", linewidth = 0.2) +
  facet_wrap(vars(mz_rt), scales = "free_y") + 
  theme_clean() 
```

We do see a decrease for both features. I would need to confirm with targeted MS/MS.


### Lyc/soy effect
2. lycopene and/or soy isoflavones effect: join a list that only keeps features that are:
- significantly different between post-Red and post-Yellow
- and significant between pre- and post-Red. 

```{r}
# combine sig features from post-red vs post-yellow
sig_postintervention_noOutlier <- full_join(postRed_sig_intrvntn_comp_clusters,
                                               postYellow_sig_intrvntn_comp_clusters)
dim(sig_postintervention_noOutlier)

# select only sig features that are present when comparing pre-Red to post-Red
lyc_soy_effect <- inner_join(sig_postintervention_noOutlier,
                             red_sig_prepost_comp_clusters,
                             by = "mz_rt") 
dim(lyc_soy_effect)
```

Export venned list
```{r}
write_csv(lyc_soy_effect,
          "Feature lists/sig-features-lyc-soy-effect.csv")
```


#### Boxplots


##### all sig tomato features

```{r}
# metabs with pval < 0.05 and fc >= 1.51
sigmetabs_lycsoy_effect <- lyc_soy_effect$mz_rt 
```

```{r}
lycsoy_effect_df_for_stats_wide <- df_for_stats %>%
  dplyr::select(!rel_abund) %>%
  pivot_wider(names_from = mz_rt,
              values_from = rel_abund_log2) %>%
  dplyr::select(c(1:11),
                sigmetabs_lycsoy_effect)

# make tidy df
lycsoy_effect_df_for_stats_tidy <- lycsoy_effect_df_for_stats_wide %>%
  pivot_longer(cols = 12:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund_log2")
```

```{r}
# changing factor levels for pre_post_intervention
lycsoy_effect_df_for_stats_tidy$pre_post_intervention <- factor(lycsoy_effect_df_for_stats_tidy$pre_post_intervention,
                              levels = c("pre_Yellow", "post_Yellow", "pre_Red", "post_Red"))

levels(lycsoy_effect_df_for_stats_tidy$pre_post_intervention)  
```

```{r, fig.height=40, fig.width=30}
(bps_lycsoy_effect <- lycsoy_effect_df_for_stats_tidy %>% 
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = c("Yellow" = "gold",
                                           "Red" = "tomato1")) +
  geom_line(aes(group = Subject), fill = "brown", linewidth = 0.2) +
  facet_wrap(vars(mz_rt), scales = "free_y") + 
  theme_clean()) 
```

Export plot
```{r, eval=FALSE}
ggsave(plot = bps_lycsoy_effect,
       filename = "plots and figures/bps_lycsoy_effect.svg",
       width = 40,
       height = 45,
       scale = 1)
```

### Low carotenoid tomato effect
2. yellow tomato effect: use list that only keeps features that are both significant between post-Red and post-Yellow timepoints, and significant between pre- and post-Yellow. These features are NOT significant between pre-Yellow and post-Yellow.

```{r}
# sig features from post-red vs post-yellow
dim(sig_postintervention_noOutlier)

# select only sig features that are present when comparing pre-Yellow to post-Yellow
low_carot_tomato_effect <- inner_join(sig_postintervention_noOutlier,
                             yel_sig_prepost_comp_clusters,
                             by = "mz_rt") 
dim(low_carot_tomato_effect)
# remove features that are significant pre-Red vs post-Red comparison
low_carot_tomato_effect <- anti_join(low_carot_tomato_effect,
                            red_sig_prepost_comp_clusters,
                            by = "mz_rt")
dim(low_carot_tomato_effect) # how many features were removed?
```

Export venned list
```{r}
write_csv(low_carot_tomato_effect,
          "Feature lists/sig-features-low-carot-tomato-effect.csv")
```



#### Boxplots


##### all sig yellow tom features

```{r}
# metabs with pval < 0.05 and fc >= 1.51
sigmetabs_yellow_effect <- as.character(low_carot_tomato_effect$mz_rt)
```

```{r}
yellow_effect_df_for_stats_wide <- df_for_stats %>%
  dplyr::select(!rel_abund) %>%
  pivot_wider(names_from = mz_rt,
              values_from = rel_abund_log2) %>%
  dplyr::select(c(1:11),
                sigmetabs_yellow_effect)

# make tidy df
yellow_effect_df_for_stats_tidy <- yellow_effect_df_for_stats_wide %>%
  pivot_longer(cols = 12:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund_log2")
```

```{r}
# changing factor levels for pre_post_intervention
yellow_effect_df_for_stats_tidy$pre_post_intervention <- factor(yellow_effect_df_for_stats_tidy$pre_post_intervention,
                              levels = c("pre_Yellow", "post_Yellow", "pre_Red", "post_Red"))

levels(yellow_effect_df_for_stats_tidy$pre_post_intervention)  
```

```{r, fig.height=25, fig.width=12}
yellow_effect_df_for_stats_tidy %>% 
  ggplot(aes(x = pre_post_intervention, y = rel_abund_log2, fill = Intervention)) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = c("Yellow" = "gold",
                                           "Red" = "tomato1")) +
  geom_line(aes(group = Subject), fill = "brown", linewidth = 0.2) +
  facet_wrap(vars(mz_rt), scales = "free_y") + 
  theme_clean() 
```




